---
title: "Example for using the CER Method for a trial design (with subgroups) with adagraph"
vignette: >
  %\VignetteIndexEntry{Example of using the CER Method for a trial design (with subgroups)}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(adagraph)
library(graphicalMCP)
```

We want to to explain how to design and test a two-stage, multi-arm, multi-endpoint clinical trial with adaptations after the interim analysis using the CER method.
We consider the design of a phase-3 clinical trial comparing two doses (high and
low) of a new compound for the treatment of severe oral mucositis (SOM) in
patients undergoing postoperative radiotherapy for head and neck cancer.
There are two endpoints being tested, the primary endpoint is to reduce the
incidence of SOM and the secondary endpoint is to reduce the duration of SOM.
Additionally, we will later want to test the effect of the treatment on the
subgroup of HPV postive patients, since it is clinically plausible that the
treatment might be more effective for this group.

# Trial without subgroup analysis

To work with the trial in `adagraph`, we first need to make a `trial_design` object with which we save all our prespecified trial specifications.

First, we define the names of our arms and endpoints, so we can recognize them
in all outputs.

```{r}
names_arms <- c("high", "low")
names_endpoints <- c("prim", "sec")
```

Then, we define everything to define a valid test procedure.

 * `test_m`, the matrix specifing the graph used for the closed testing procedure. The matrix we want to use is depicted in @fig-graph
 * `weights`, the corresponding weights for the closed testing procedure
 * `alpha`, the overall FWER we want to control for
 * `alpha_spending_function`, for specifing how much of the FWER should be spent at the interim test
 * `t`, the time information fraction at which the interim test is planned. Since for our trial, it is planned that the interim look is performed after 50% of the participants have enrolled, t is 1/2.

```{r}
test_m <- rbind(
  c(0, 0.6, 0.4, 0),
  c(0.6, 0, 0, 0.4),
  c(0, 1, 0, 0),
  c(1, 0, 0, 0)
)
weights <- c(0.5, 0.5, 0, 0)

as <- function(x, t) 2 - 2 * stats::pnorm(stats::qnorm(1 - x / 2) / sqrt(t))
alpha <- 0.025
t <- 1 / 2
```

To be able to use the correlation structure between the endpoints, we also need
to specify the patient numbers for each group.
There are two ways to do this in our case. First, we can define the case numbers
in a dataframe, noting for each arm the number of patients we will recruit.
The patient numbers are then only the amount of patients expected (or already
realised) for the first stage of the trial, 35 per arm in our case.
We generally assume that the ratios between the different groups stay the same,
but variations from this, either by changes in the strucutre or from random
fluctionations, can be given as adaptations later on.
The relative size of the first and second stage is given via the information
fraction argument `t`.

```{r}
n_table <- data.frame(
  arm = c("control", "high", "low"),
  n = c(35, 35, 35)
)
```

Since there are no subgroups, we can alternatively also use simple numeric
vectors for the control and arms separately.
(Note that in this case the names in the vector entry are not necessary, only
the order is relevant.)

```{r}
n_control <- 35
n_arms <- c(high = 35, low = 35)
```

By default, we also use the fact that if $B_J(\chi_1)\geq 1$ for some intersection hypotheses, we can also reject that subhypotheses at the interim, as described in the beginning of section 3.2.3.
If this is not the desired behaviour, set `seq_bonf` to `FALSE`.
We now have all we need to define our testing design.

```{r}
design <- trial_design(
  arm = 2,
  endpoints = 2,
  n_control = n_control,
  n_arms = n_arms,
  weights = weights,
  t = t,
  alpha = alpha,
  test_m = test_m,
  names_arms = names_arms,
  names_endpoints = names_endpoints,
  alpha_spending_f = as,
  seq_bonf = TRUE
)

design
```

As we can see, the print method gives a short description of the design.
For a more detailed overview, showing all arguments used, we can use the
`summary()` function.

```{r}
summary(design)
```

Additonally, we can export and print the graph with the help of the
`graphicalMCP`
package.

```{r}
#| label: fig-graph
#| fig-cap: Graph for the testing procedure we are using, generated using the
#|   `export_graphical_mcp()` function
gmcp_obj <- design |> export_graphical_mcp() |> plot(nrow = 2)
```

Assume now the we get after the unblinding for the interim analysis the p values are $p_{1,1} = 0.00445$, $p_{2,1} = 0.0952$, $p_{3,1} = 0.0225$, and $p_{4,1} = 0.1104$.
We incorporate this information by applying the `cer_interim_test` function.
The result is again a `trial_design` object, which now includes the interim test
results.

```{r}
design_interim <- cer_interim_test(
  design,
  p_values = c(0.00045, 0.0952, 0.0225, 0.1104)
)

design_interim
```

We can see that for now, only the Hypothesis testing the primary endpoint for
the low dose is rejected.
While it is possible to continue now without an adaptations to the design,
`adagraph` allows for many changes to be made to the 
Adagraph does by itself not assume that we will change anything about our trial design, even if a hypotheses was rejected.

Assume that it is decided that for the high dose, since the primary hypothesis
was already rejected, it is not necessary to continue recruitment and the focus
will instead be on the low dose.
For dropping a specific arm, we can use the `trial_drop_arms()` function, which
is made exactly for this purpose.
This change has the benefit of allowing more patients for the low dose and control to
be recruited than previously planned.
This results in a change to the `t` parameter and the correlation structure, which can be calculated automatically with the `trial_adapt_n` function.
Additionally, the weights will be redistributed equally on the two
remaining hypotheses.
For adjusting the weights, we use the much more general `cer_adapt()` function.

```{r}
n_control_ad <- 53
n_arms_ad <- c(0, 52) #even tho it is being dropped, we still need to give
#n for the low dose arm

design_adj <- design_interim |>
  trial_drop_arms("high") |>
  trial_adapt_n(
    n_control_2 = n_control_ad,
    n_arms_2 = n_arms_ad
  ) |>
  cer_adapt(
    weights = c(0, 0.5, 0, 0.5),
  )

design_adj
```

Note the new testing matrix resulting from those changes:

```{r}
design_adj$ad_test_m
```

Now we can test for the final results, $\tilde{p}_{2,2} = 0.0111$, and $\tilde{p}_{4,2}$.
By default `adagraph` expects the already combined p values, but using the
`combined = FALSE` option adagraph calculates from the second stage p values the
combined values, using the adapted value for `t` (which was calculated
automatically by the `trial_adapt_n()` function).
Note that the hypthoses that we no further use are still present in the design object, they just don't get any weight assigned.
Therefore, we still need to use a vector of length 4 for our p-values, but can use `NA` as a corresponding value.


```{r}
design_tested <- cer_final_test(design_adj, c(NA, 0.0111, NA, 0.0234))

design_tested
```

As we can see, both the hypotheses that we are still testing for are now also being rejected.

# Enrichtment design

Take now the same trial as before, but assuming a bigger sample size of 150
people per arm.
Thanks to the resulting higher power, it is decided to also check for effects
specifically in the subgroup of patients testing positive for the human
papilloma virus (HPV+).
Doing this is also possible with `adagraph`, using again the `trial_design`
class.

Again, we define the names of our arms, endpoints and subgroups, as well as the
characteristics of the design.
This is now a bit more complicated, since we have a total of 8 hypotheses, so we
use some helper variables.
```{r}
names_arms <- c("high", "low")
names_endpoints <- c("prim", "sec")
names_subgroups <- "HPV+"


alpha <- 0.025
t <- 0.5
weights <- c(0.35, 0.35, 0, 0, 0.15, 0.15, 0, 0)

s_p <- 1 / 3 #secondary to primary
p_p <- 0.2 #primary to primary
p_s <- 0.4 #primary to secondary

#fmt: skip
test_m <- matrix(
    byrow = TRUE,
    ncol = 8,
    data = c(
      0, p_p, p_s,   0, p_p, p_p,   0,   0,
    p_p,   0,   0, p_s, p_p, p_p,   0,   0,
      0, s_p,   0,   0, s_p, s_p,   0,   0,
    s_p,   0,   0,   0, s_p, s_p,   0,   0,
    p_p, p_p,   0,   0,   0, p_p, p_s,   0,
    p_p, p_p,   0,   0, p_p,   0,   0, p_s,
    s_p, s_p,   0,   0,   0, s_p,   0,   0,
    s_p, s_p,   0,   0, s_p,   0,   0,   0
  ))
```

For defining the first stage sample sizes, we need to use the `n_table` data
frame, where we also include the status of each group for each row (which is
only `HPV+` in our case).

```{r}
n_table <- rbind(
  data.frame(arm = "control", `HPV+` = FALSE, n = 20, check.names = FALSE),
  data.frame(arm = "control", `HPV+` = TRUE, n = 15, check.names = FALSE),
  data.frame(arm = "high", `HPV+` = FALSE, n = 20, check.names = FALSE),
  data.frame(arm = "high", `HPV+` = TRUE, n = 15, check.names = FALSE),
  data.frame(arm = "low", `HPV+` = FALSE, n = 20, check.names = FALSE),
  data.frame(arm = "low", `HPV+` = TRUE, n = 15, check.names = FALSE)
)

design <- trial_design(
  arm = 2,
  endpoints = 2,
  subgroups = 1,
  n_table = n_table,
  weights = weights,
  t = t,
  alpha = alpha,
  test_m = test_m,
  names_arms = names_arms,
  names_endpoints = names_endpoints,
  names_subgroups = names_subgroups,
  alpha_spending_f = as,
  seq_bonf = TRUE
)

design
```

Again we get a detailed summary when using the `summary()` function.

```{r}
summary(design)
```


```{r}
#| label: fig-graph-subgroup
#| fig-cap: Graph for the testing procedure we are using, generated using the
#|   `export_graphical_mcp()` function
gmcp_obj <- design |> export_graphical_mcp() |> plot(nrow = 4)
```

For the example, we assume that for the total group, we get $p_{1,1} = 0.00445$, $p_{2,1} = 0.0952$, $p_{3,1} = 0.0225$, and $p_{4,1} = 0.1104$.
Additionally, there are p values for only the subgroup of the HPV positive
popultaion, where we have $p_{5,1} = 0.0532$, $p_{6,1} = 0.2152$, $p_{7,1} = 0.0352$, and $p_{8,1} = 0.1728$.
We incorporate this information by applying the `cer_interim_test` function.
The result is again a `trial_design` object, which now includes the interim test
results.

```{r}
design_interim <- cer_interim_test(
  design,
  p_values = c(0.00045, 0.0952, 0.0225, 0.1104, 0.0532, 0.2152, 0.0352, 0.1728)
)

design_interim
```

The further use of the package proceeds as above.
One important difference is that we have to continue using the n_table
format, since there are still subgroups present.
It is however possible to emit rows that are empty.
There is now of course the option to also drop a whole subgroup from the
analysis.
For example, to arrive at the same case as before, we use the following code.

```{r}
n_table <- rbind(
  data.frame(arm = "control", `HPV+` = FALSE, n = 53, check.names = FALSE),
  data.frame(arm = "low", `HPV+` = FALSE, n = 52, check.names = FALSE)
)

design_adj <- design_interim |>
  trial_drop_arms("high") |>
  trial_drop_groups("HPV+") |>
  trial_adapt_n(
    n_control_2 = n_control_ad,
    n_arms_2 = n_arms_ad
  ) |>
  cer_adapt(
    weights = c(0, 0.5, 0, 0.5, 0, 0, 0, 0),
  )

design_adj
```

This leads then to the same result as before.

```{r}
design_tested <- cer_final_test(
  design_adj,
  c(NA, 0.0111, NA, 0.0234, NA, NA, NA, NA)
)

design_tested
```
